\chapter*{Master's Thesis Proposal}

\begin{tabular}{lp{10.1cm}}
		\hline
		\textbf{Author} &\href{mailto:\Email}{\AutorDP}\\
		\textbf{Supervisor} &\href{mailto:\EmailSup}{\Supervisor}\\
		\textbf{Proposed topic} &\Bookname\\
		\hline
\end{tabular}

\bigskip

\small
\paragraph{Motivation}
XXX



\paragraph{Hypotheses}
\begin{enumerate}
		\item[] XXX xxx
\end{enumerate}

\paragraph{Methodology}


\paragraph{Expected Contribution}


\paragraph{Outline}
\begin{enumerate}
	\item Introduction: Explains thesis reasoning
	\item Analysis of Flow Cytometry Data: Summary of relevant knowledge regarding flow cytometry and data analysis
	\item Automatic clustering: 
	\item DESCRIPTION OF MY WORK: find a different name for this section 
	\item Discussion and limitations: Thesis limitations are addressed
	\item Concluding remarks: Findings and implications summary
\end{enumerate}


\paragraph{Core bibliography}


\begin{enumerate}
\item[]S. Das, B. Saket, B. C. Kwon, \& A. Endert, "Geono-Cluster: Interactive Visual Cluster Analysis for Biologists," in IEEE Transactions on Visualization and Computer Graphics, vol. 27, no. 12, pp. 4401-4412, 1 Dec. 2021, doi: 10.1109/TVCG.2020.3002166.
\item[]Sieger, T., Hurley, C. B., Fišer, K., \& Beleites, C. (2017). Interactive Dendrograms: The R Packages idendro and idendr0. Journal of Statistical Software, 76(10), 1–22.
\item[]Šmelko, A., Kratochvíl, M., Kruliš, M., \& Sieger, T. (2021, September). GPU-Accelerated Mahalanobis-Average Hierarchical Clustering Analysis. In European Conference on Parallel Processing (pp. 580-595). Springer, Cham.
\item[]Fišer, K., Sieger, T., Schumich, A., Wood, B., Irving, J., Mejstříková, E., \& Dworzak, M. N. (2012). Detection and monitoring of normal and leukemic cell populations with hierarchical clustering of flow cytometry data. Cytometry Part A, 81(1), 25-34.

\end{enumerate}


\vfill
\begin{table}[!hbp]
\begin{tabular}{lr}

 \begin{tabular}{p{3.5cm}}
     \hline \hspace{1cm} Author
 \end{tabular}
 
 \hspace{5.5cm}
 
 \begin{tabular}{p{3.5cm}}
     \hline \hspace{0.8cm} Supervisor
 \end{tabular}

 
 \end{tabular}
 \end{table}

\normalsize





